Persistence in functional topology: q-tameness, local-connectivity and a correction to a theorem of Morse

### 2021-04-01 connectivity.tex
Split preliminaries

\begin{defi} \label{defi:local_connectedness}
	For $n \in \Z$ a continuous map is said to be \textit{$n$-homologically small} or \textit{trivial}, denoted $\HS_n$ and $\HT_n$ respectively, if the image of the map induced by $\tilde H_n(-)$ is finitely generated or 0, and we say it is \textit{homologically small} or \textit{trivial}, denoted respectively $\HS$ and $\HT$, if it is $\HS_n$ or $\HT_n$ for every~$n$.

	A space $X$ is said to be \emph{homologically locally connected}, denoted $\HLC$, if for each $x \in X$ any neighborhood $V$ of $x$ contains a neighborhood $U$ of $x$ such that the inclusion $U \to V$ is $\HT$.
\end{defi}

\anibal{Use a different example, $\H$ not yet defined.}
\begin{ex}
	The $d$-dimensional Hawaiian earring $\H$ is not $\HLC$. However, if we consider $\H$ as a subspace of $\mathbb{H}^{d'}$ via the obvious embedding for $d < d'$, then $\H$ is $\HLC$ in $\mathbb{H}^{d'}$.
\end{ex}

Local connectivity assumptions have a long history of being used to ensure that different homology theories agree and to ensure that homology is finite dimensional, see \cite{MR0007094} for early examples. For us, the following is relevant.

@@

\begin{cor}\label{cor:cech_sing_persistent_iso}
	Let $M$ be a metric space and $f \colon M \to \mathbb{R}$ a function such that $M_{\leq c}$ is a paracompact $\HLC$-space for each $c \in \mathbb{R}$.
	Then $\CH_n(M_{\leq \bullet}) \cong H_n(M_{\leq\bullet})$ for any integer $n$.
\end{cor}

For more results of this type, we refer to \cite{MR1481706}. There, the above result \cite[Corollary VI.12.6]{MR1481706} and many similar ones are shown using sheaf theoretic methods. For example, Bredon also discusses cohomology local connectedness \cite[Section II.17]{MR1481706} and various other comparison results, e.g.\@ between singular and Borel-Moore homology \cite[Corollary V.12.15]{MR1481706} in the presence of homology local connectedness, as well as examples such as spaces that are sheaf cohomology locally connected but not homology locally connected \cite[Example II.17.12]{MR1481706}. Using Bredon's results, we also get the following.

\anibal{Does this needs field coefficients? (}
\begin{prop} \label{prop:fin_dim_sing_hom}
	Let $X$ be compact Hausdorff and $\HLC$.
	Then $H_n(X; \mathbb{F})$ is finite-dimensional for all $n$ and any field $\mathbb{F}$.
\end{prop}

\begin{proof}
	By the universal coefficient theorem for homology, trivial singular homology with integer coefficients also implies trivial singular homology for any other coefficient group. In particular, this means that $X$ being $HLC^{\infty}$ also implies that $X$ is $HLC^{\infty}_{\mathbb{F}}$. By the universal coefficient theorem for cohomology, the singular cohomology with coefficients in $\mathbb{F}$ is naturally isomorphic to the dual space of the singular homology with coefficients in $\mathbb{F}$. This means that being $HLC^{\infty}_{\mathbb{F}}$ implies being locally connected with respect to singular cohomology in all degrees with coefficients in $\mathbb{F}$. Since we assume $X$ to be $HLC^{\infty}$, \cite[Theorem III.1.1]{MR1481706} now implies that $X$ is locally connected with respect to sheaf cohomology with coefficients in the locally constant sheaf with value $\mathbb{F}$. \cite[Corollary II.17.7]{MR1481706} implies that the sheaf cohomology of the whole space with coefficients in this sheaf is finite-dimensional, so again by \cite[Theorem III.1.1]{MR1481706} the same holds for singular cohomology with coefficients in $\mathbb{F}$. Applying the universal coefficient theorem once more proves the claim.
\end{proof}

@@

\begin{cor}
	Let $X$ be a topological space and $f \colon X \to \R$ a function such that $f_{\leq c}$ is a compact Hausdorff $\HLC$-space for each $c \in \R$.
	Then $H_d(f_{\leq\bullet})$ is p.f.d. for any $d$.
\end{cor}

\begin{defi}
	Let $M$ be a metric space and $f \colon M \to \R$ a function.
	The space $M$ is said to be \textit{locally-$f$-connected} if for each $p \in M$, $e > 0$, and $n \in \N$ there exists $\delta \in (0, e)$ such that for every $c \geq 0$ the inclusion $B_\delta(p) \cap M_{\leq f(p)+\delta+c} \to B_e(p) \cap M_{\leq f(p)+e+c}$ is $\HT$.
\end{defi}

>>>>

Let $\H$ be a fixed homology theory.

\begin{defi} \label{defi:local_connectedness}
	For $n \in \Z$ a continuous map is said to be \textit{$n$-homologically small} or \textit{trivial}, denoted $\HS_n$ and $\HT_n$ respectively, if the image of the map induced by $\H_n$ is finitely generated or 0, and we say it is \textit{homologically small} or \textit{trivial}, denoted respectively $\HS$ and $\HT$, if it is $\HS_n$ or $\HT_n$ for every~$n$.
\end{defi}

\begin{defi}
	A space $X$ is said to be \emph{homologically locally connected}, denoted $\HLC$, if for each $x \in X$ any neighborhood $V$ of $x$ contains a neighborhood $U$ of $x$ such that the inclusion $U \to V$ is $\HT$.

	A sublevel filtration $(X,f)$ is said to be $\emph{homologically locally connected}$, denoted $\HLC$, if for each $x \in X$, neighborhood $V$ of $x$ and $\epsilon > 0$ there exists $U$ a neighborhood of $x$ with $U \subseteq V$ and $\delta \in (0, \epsilon)$ such that
	\begin{equation*}
	X_{\leq f(x) + \delta} \cap U \to X_{f(x) + \epsilon} \cap V
	\end{equation*}
	is $\HT$.
\end{defi}

%\anibal{Use a different example, $\HE$ not yet defined.}
%\begin{ex}
%	The $d$-dimensional Hawaiian earring $\HE$ is not $\HLC$. However, if we consider $\HE$ as a subspace of $\mathbb{H}^{d'}$ via the obvious embedding for $d < d'$, then $\HE$ is $\HLC$ in $\mathbb{H}^{d'}$.
%\end{ex}

Local connectivity assumptions have a long history of being used to ensure that different homology theories agree and to ensure that homology is finite dimensional, see \cite{MR0007094} for early examples.
For us, the following is relevant.
Recall that a space is said to be \textit{paracompact} if every open cover has a locally finite open refinement, i.e., every point has a neighborhood intersecting finitely many sets in the refinement.

@@

\begin{cor}\label{cor:cech_sing_persistent_iso}
	Let $(X, f)$ be a sublevel filtration. If each sublevel set is paracompact and $\HLC$, then its singular and \v{C}ech persistence modules coincide.
\end{cor}

For more results of this type, we refer to \cite{MR1481706}. There, the above proposition \cite[Corollary VI.12.6]{MR1481706} and many similar ones are proven using sheaf theoretic methods.
For example, Bredon also discusses cohomology local connectedness \cite[Section II.17]{MR1481706} and various other comparison results, e.g.\@ between singular and Borel-Moore homology \cite[Corollary V.12.15]{MR1481706} in the presence of homology local connectedness, as well as examples such as spaces that are sheaf cohomology locally connected but not homology locally connected \cite[Example II.17.12]{MR1481706}.
Using Bredon's results, we also get the following.

\begin{prop} \label{prop:fin_dim_sing_hom}
	Let $X$ be a compact Hausdorff $\HLC$ space.
	Then the singular homology of $X$ with field coefficients is finite-dimensional in every degree.
\end{prop}
\anibal{Does this really needs field coefficients?} 
\begin{proof}
	By the universal coefficient theorem for homology, trivial singular homology with integer coefficients also implies trivial singular homology for any other coefficient group. In particular, this means that $X$ being $HLC^{\infty}$ also implies that $X$ is $HLC^{\infty}_{\mathbb{F}}$. By the universal coefficient theorem for cohomology, the singular cohomology with coefficients in $\mathbb{F}$ is naturally isomorphic to the dual space of the singular homology with coefficients in $\mathbb{F}$. This means that being $HLC^{\infty}_{\mathbb{F}}$ implies being locally connected with respect to singular cohomology in all degrees with coefficients in $\mathbb{F}$. Since we assume $X$ to be $HLC^{\infty}$, \cite[Theorem III.1.1]{MR1481706} now implies that $X$ is locally connected with respect to sheaf cohomology with coefficients in the locally constant sheaf with value $\mathbb{F}$. \cite[Corollary II.17.7]{MR1481706} implies that the sheaf cohomology of the whole space with coefficients in this sheaf is finite-dimensional, so again by \cite[Theorem III.1.1]{MR1481706} the same holds for singular cohomology with coefficients in $\mathbb{F}$. Applying the universal coefficient theorem once more proves the claim.
\end{proof}

@@

\begin{cor}
	Let $(X, f)$ be a sublevel filtration. If each sublevel set is compact and $\HLC$
	Then its singular persistence module is p.f.d. in every degree.
\end{cor}

### 2021-04-01 counterexample.tex
Changes before meeting
Split preliminaries

\section{Counterexample}

\begin{figure}[t]
	\centering
	\includegraphics[width=.77\textheight]{hawaiian_earring}
	\hfill
	\caption{TBW}
	\label{fig:earrings}
\end{figure}

In this section we show that weak local-$f$-connectivity is not sufficient to ensure q-tameness and, in the next, that local-$f$-connectivity is.

The $d$-dimensional Hawaiian earring is defined as the subspace
\begin{equation*}
\H = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\mathbb{R}^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}
\end{equation*}
of $d+1$-dimensional Euclidean space.
Let $f \colon \mathbb{H}^{d}\to\mathbb{R}$ as $0$ at the origin $\vec{0}$ and $1$ everywhere else.
All sublevel sets $\H_{\leq t}$ are either the singleton $\{\vec{0}\}$ or the whole space $\H$, and therefore compact.
Moreover, $\mathbb{H}^{d}$ is locally-$f$-connected as we prove now.

Let $e > 0$ and $p \in \H$.
If $p$ is the origin, we choose $\delta$ between 0 and $\min\{e,1\}$.
Then $\H_{\leq f(p) + \delta} = \H_{\leq \delta} = \{p\}$, so the condition is trivially satisfied.
For $p$ not the origin, there is a unique $d$-sphere that contains it.
Clearly, we may choose $\delta$ so small that $B_{\delta}(p)$ contained in this sphere, so $B_{\delta}(p)$ is a disk and the condition follows trivially again.

What remains to be shown is that $\H_{\leq \bullet}$ is not q-tame with respect to singular and \v{C}ech homology.
Note that $\H_{\leq t}$ is constant with value $\H$ for $t \geq 1$, so it suffices to show that the singular and \v{C}ech homology of $\H$ with field coefficients is infinite dimensional.
The case of singular homology is well know.
For \v{C}ech homology we use the fact that it commutes with inverse limits for compact Hausdorff spaces.
Define 
\begin{align*}
\mathbb{H}^{d}_k&=\left\{(x_0,\dots,x_d)\in\mathbb{R}^{d+1}\mid \left(x_0-\frac{1}{k}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{k}\right)^2\right\}\\
&\cup\bigcup_{n=1}^{k-1}\left\{(x_0,\dots,x_d)\in\mathbb{R}^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k}\mathbb{H}^{d}_{k}=\bigcap_{k}\mathbb{H}^{d}_{k}=\mathbb{H}^{d}$, and hence $\CH_{d}(\mathbb{H}^{d}) = \lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})$.
One can easily check that each $\mathbb{H}^{d}_{k}$ satisfies the assumptions for \cref{prop:cech_sing_hom_hlc}, which implies $\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})=\lim_{k}H_{d}(\mathbb{H}^{d}_{k})$.
We compute
\[
\lim_{k}H_{d}(\mathbb{H}^{d}_{k})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
\]
which is infinite-dimensional over $\mathbb{F}$.
This finishes the proof.

>>>>

\section{Counterexample} \label{sec:counterexample}

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale=88]
	\draw[thick] (-.1,0) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05); % remove for all circles
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{TBW}
	\label{fig:earrings}
\end{figure}

In this section we show that a sublevel filtration can be $\HLC$ and not q-tame.

The $d$-dimensional Hawaiian earring is defined as the subspace
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}
\end{equation*}
of $d+1$-dimensional Euclidean space.
Let $f \colon \HE\to\R$ as $0$ at the origin $\vec{0}$ and $1$ everywhere else.
All sublevel sets $\HE_{\leq t}$ are either the singleton $\{\vec{0}\}$ or the whole space $\HE$, and therefore compact.
Moreover, $\HE$ is locally-$f$-connected as we prove now.

Let $e > 0$ and $p \in \HE$.
If $p$ is the origin, we choose $\delta$ between 0 and $\min\{e,1\}$.
Then $\HE_{\leq f(p) + \delta} = \HE_{\leq \delta} = \{p\}$, so the condition is trivially satisfied.
For $p$ not the origin, there is a unique $d$-sphere that contains it.
Clearly, we may choose $\delta$ so small that $B_{\delta}(p)$ contained in this sphere, so $B_{\delta}(p)$ is a disk and the condition follows trivially again.

What remains to be shown is that $\HE_{\leq \bullet}$ is not q-tame with respect to singular and \v{C}ech homology.
Note that $\HE_{\leq t}$ is constant with value $\HE$ for $t \geq 1$, so it suffices to show that the singular and \v{C}ech homology of $\HE$ with field coefficients is infinite dimensional.
The case of singular homology is well know.
For \v{C}ech homology we use the fact that it commutes with inverse limits for compact Hausdorff spaces.
Define 
\begin{align*}
\HE_k &= \left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{k}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{k}\right)^2\right\}\\
&\cup\bigcup_{n=1}^{k-1}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k}\HE_{k} = \bigcap_{k}\mathbb{H}^{d}_{k}=\mathbb{H}^{d}$, and hence $\CH_{d}(\HE) = \lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})$.
One can easily check that each $\mathbb{H}^{d}_{k}$ satisfies the assumptions for \cref{prop:cech_sing_hom_hlc}, which implies $\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})=\lim_{k}H_{d}(\mathbb{H}^{d}_{k})$.
We compute
\[
\lim_{k}H_{d}(\mathbb{H}^{d}_{k})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
\]
which is infinite-dimensional over $\mathbb{F}$.
This finishes the proof.

### 2021-04-01 persistence.tex
Split preliminaries

\section{Preliminaries}

\subsection*{Notation and Conventions}
For the rest of the paper, we let $\mathbb{F}$ be a field and whenever we write homology without explicit mention of coefficients, we mean homology with coefficients in $\mathbb{F}$.

\subsection{Sublevel set filtrations}

Treat in parallel the case of spaces $X$ and the case of sublevel filtrations $(X, f)$. Leads nicely to different connectivities.
\begin{itemize}
	\item metric space
	\item upper semi-continuous function
	\item sublevel set filtrations
\end{itemize}

\begin{defi}
	Let $M$ be a metric space. For any $p \in M$ and $\epsilon \geq 0$ denote
	\begin{equation*}
	B_\epsilon(p) = \{q \in M\ |\ d(p,q) < \epsilon\}, \qquad
	\overline B_\epsilon(p) = \{q \in M\ |\ d(p,q) \leq \epsilon\}.
	\end{equation*}
\end{defi}

\begin{defi}
	Let $X$ be a space and $f \colon X \to \R$ a function. For any $t \in \R$, the \textit{$t$-sublevel set of $f$} is defined by 
	\begin{equation*}
	X_{\leq t} = f^{-1}((-\infty, t]).
	\end{equation*}
\end{defi}

\subsection{Persistence homology}

\begin{itemize}
	\item Persistence module
	\item q-tameness
	\item Barcode
	\item bottleneck
	\item stability
\end{itemize}

>>>>

\subsection{Persistence}

\begin{itemize}
	\item Persistence module
	\item p.f.d
	\item q-tameness
	\item Barcode
	\item bottleneck
	\item stability
\end{itemize}

### 2021-04-01 theorem.tex
Split preliminaries

@@

\begin{lem} \label{l:key lemma for q-tameness}
	Let $M$ and $f$ be as in Theorem \ref{t:strong local connectenss implies q-tameness}.
	Consider sets $K, L \subseteq M$ with $K$ compact and $K \subseteq \interior(L)$. For any $s < t$ there is $\delta \in (0,\, t-s)$ such that $K \cap M_{s+\delta} \to L \cap M_{t}$ is $\HS$.
\end{lem}

@@

	We will now show that the class $\Sigma_{s,t}$ is closed under finite unions.
	For $i \in \{1, 2\}$ let $K_i$ be in $\Sigma_{s,t}$ with $\delta_i > 0$ and $K_i \subseteq U_i$ open such that $U_{i} \cap M_{s+\delta_i} \to L \cap M_{t}$ is $\HS_n$.
	We Use Lemma \ref{l:neighborhood third} to construct sets $K_i^\prime$ such that
	\begin{equation*}
	K_i \subseteq \interior(K_i^\prime) \subseteq K_i^\prime \subseteq U_i.
	\end{equation*}
	Notice that for $\delta = \min(\delta_i)$ we have $K_i^\prime \cap M_{s+\delta} \to L \cap M_t$ is $\HS_n$.
	Additionally, the induction hypothesis implies that $K_1 \cap K_2 \cap M_s \to K_1^\prime \cap K_2^\prime \cap M_{s+\delta}$ is $\HS_{(n-1)}$.
	We therefore have the following commutative diagram satisfying the assumptions of Lemma~\ref{l:commutative algebra}:
	\begin{equation*}
	\begin{tikzcd}
	H_n(L \cap M_t) \oplus H_n(L \cap M_t) \arrow[r] &
	H_n(L \cap M_t) & \\
	H_{n}(K_1^\prime \cap M_{s+\delta}) \oplus H_n(K_2^\prime \cap M_{s+\delta}) \arrow[r] \arrow[u] & 
	H_{n}((K_1^\prime \cap M_{s+\delta}) \cup (K_2^\prime \cap M_{s+\delta})) \arrow[r] \arrow[u] &
	H_{n-1}(K_1^\prime \cap K_2^\prime \cap M_{s+\delta}) \\ & 
	H_{n}((K_1 \cup K_2) \cap M_s) \arrow[r] \arrow[u] &
	H_{n-1}(K_1 \cap K_2 \cap M_s). \arrow[u]
	\end{tikzcd}
	\end{equation*}
	We conclude that $K_1 \cup K_2 \in \Sigma_{s, t}$.
	Since any compact $K \subseteq \interior(L)$ can be expressed as a finite union of sets in $\Sigma_{s,t}$ the induction step and the lemma are proven.
\end{proof}

>>>>

In this section we introduce a stronger version of $\HLC$ for sublevel filtrations, and we show that Morse filtrations with this property are q-tame.

\begin{defi}
	A sublevel filtration $(X,f)$ is said to be \textit{strongly} $\HLC$ if for each $x \in X$, neighborhood $V$ of $x$ and $\epsilon > 0$ there exists $U$ a neighborhood of $x$ with $U \subseteq V$ and $\delta \in (0, \epsilon)$ such that
	\begin{equation*}
	X_{\leq f(x) + \delta + c} \cap U \to X_{f(x) + \epsilon + c} \cap V
	\end{equation*}
	is $\HT$ for every $c \geq 0$.
\end{defi}

@@

\begin{lem} \label{l:key lemma for q-tameness}
	Fix a homology theory. Let $M$ and $f$ be as in Theorem \ref{t:strong local connectenss implies q-tameness}.
	Consider sets $K, L \subseteq M$ with $K$ compact and $K \subseteq \interior(L)$. For any $s < t$ there is $\delta \in (0,\, t-s)$ such that $K \cap M_{s+\delta} \to L \cap M_{t}$ is $\HS$.
\end{lem}

@@

	We will now show that the class $\Sigma_{s,t}$ is closed under finite unions.
	For $i \in \{1, 2\}$ let $K_i$ be in $\Sigma_{s,t}$ with $\delta_i > 0$ and $K_i \subseteq U_i$ open such that $U_{i} \cap M_{s+\delta_i} \to L \cap M_{t}$ is $\HS_n$.
	We Use Lemma \ref{l:neighborhood third} to construct sets $K_i^\prime$ such that
	\begin{equation*}
	K_i \subseteq \interior(K_i^\prime) \subseteq K_i^\prime \subseteq U_i.
	\end{equation*}
	Notice that for $\delta = \min(\delta_i)$ we have $K_i^\prime \cap M_{s+\delta} \to L \cap M_t$ is $\HS_n$.
	Additionally, the induction hypothesis implies that $K_1 \cap K_2 \cap M_s \to K_1^\prime \cap K_2^\prime \cap M_{s+\delta}$ is $\HS_{(n-1)}$.
	We therefore have the following commutative diagram satisfying the assumptions of Lemma~\ref{l:commutative algebra}:
	\begin{equation*}
	\begin{tikzcd}
	\H_n(L \cap M_t) \oplus \H_n(L \cap M_t) \arrow[r] &
	\H_n(L \cap M_t) & \\
	\H_{n}(K_1^\prime \cap M_{s+\delta}) \oplus \H_n(K_2^\prime \cap M_{s+\delta}) \arrow[r] \arrow[u] & 
	\H_{n}((K_1^\prime \cap M_{s+\delta}) \cup (K_2^\prime \cap M_{s+\delta})) \arrow[r] \arrow[u] &
	\H_{n-1}(K_1^\prime \cap K_2^\prime \cap M_{s+\delta}) \\ & 
	\H_{n}((K_1 \cup K_2) \cap M_s) \arrow[r] \arrow[u] &
	\H_{n-1}(K_1 \cap K_2 \cap M_s). \arrow[u]
	\end{tikzcd}
	\end{equation*}
	We conclude that $K_1 \cup K_2 \in \Sigma_{s, t}$.
	Since any compact $K \subseteq \interior(L)$ can be expressed as a finite union of sets in $\Sigma_{s,t}$ the induction step and the lemma are proven.
\end{proof}
